---
title: "Analysis on Physical Characteristics of Penguins in the Palmer Archipelago"
author: "Jyotika Sharma, Selena Cooper, Grace Kolker"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The goal of this project is to explore the Palmer Penguins dataset to better understand the differences in physical traits between penguin species. This dataset includes measurements collected on three penguins species (Adelie, Gentoo, and Chinstrap) living on multiple islands in the Palmer Archipelago in Antarctica. By analyzing these physical measurements, we hope to identify patterns, species differneces, and relationships between body size variables. Understanding these traits can help explain how environmental conditions and biological factors may influence species variation and survival.

In pursuit of the stated goal, we will explore the following questions:

1. How do body size measurements vary between penguin species? Are certain species consistently larger in one or more measurements such as body mass or flipper length? - Selena

2. How do bill length and bill depth interact to help differentiate penguin species? Can these measurements be used to identify species differences effectively? - Selena

3. Is there a relationship between flipper length and body mass across species? Do larger flippers indicate heavier penguins, and does this relationship differ by species? - Jyotika

4. How does sex influence body size? Are male penguins consistently larger than female penguins across all species, and does this vary in strength across difference species? - Jyotika 

5. Does the island a penguins inhabits affect its body measurements? Are physical differences across islands still present even after accounting for species? - Grace

6. Which variable (bill length, bill depth, flipper length, or body mass) provides the strongest prediction of species classification? How do combines variables compare to single-variable predication? - Grace 

These are the main questions we intend to answer through analysis of the Palmer Penguins dataset. With the findings, we will be able to draw meaningful conclusions about physical trait differences and the ecological factors that influence species variation in Antarctic penguins.

## Data

### Structure

The link to the dataset is
```{r}
library(palmerpenguins)
data(package = "palmerpenguins")
```

### Cleaning
```{r}
library(tidyverse)

penguins_clean <- penguins |> 
  drop_na()

# Check dimensions before/after
dim(penguins)
dim(penguins_clean)

```
The Palmer Penguins dataset was already very clean and well-structured upon import, containing properly formatted variable types and no obvious outliers or inconsistent entries. We did not need to convert sex to factor since it already is well. The only issues present were several missing values in the bill measurements, flipper length, body mass, and sex variables. To ensure accurate and consistent analysis across all variables, we removed rows containing any missing values using drop_na(). This reduced the dataset from 344 observations to 333 complete cases.

### Variables
```{r}
dim(penguins)
str(penguins)
```
There are 344 penguins in the dataset. There are 3 different species of penguins collected from 3 different islands in the Palmer Archipelago.

* species: the biological species of each penguin (Adelie, Gentoo, and Chinstrap)
* island: The island within the Palmer Archipelago where the penguin was observed (Biscoe, Dream, and Torgersen)
* bill_length_mm: The length of the penguin's bill (from tip to base) measured in milimeters
* bill_depth_mm: Thickness (depth) fo the penguin's bill measured at the base in milimeters
* flipper_length_mm: Length of the penguin's flipper from shoulder to tip measured in milimeters
* body_mass_g: Body weight of the penguin measured in grams
* sex: Sex of the penguin (male or female)
* year: The year the penguin was sampled during the Palmer Station field seasons (2007, 2008, 2009)

## Results

###  How do body size measurements vary between penguin species? Are certain species consistently larger in one or more measurements such as body mass or flipper length?

```{r}
ggplot(penguins, aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot() +
  labs(title = "Body Mass by Penguin Species",
       y = "Body Mass (g)") +
  theme_minimal()

ggplot(penguins, aes(x = species, y = flipper_length_mm, fill = species)) +
  geom_boxplot() +
  labs(title = "Flipper Length by Penguin Species",
       y = "Flipper Length (mm)") +
  theme_minimal()

penguins %>%
  group_by(species) %>%
  summarize(
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    mean_flipper = mean(flipper_length_mm, na.rm = TRUE)
  )

```

The plots and summary statistics together show clear, consistent differences in body size among the three penguin species. Gentoo penguins stand out as the largest across both measured traits. Their body masses cluster around 5000 g, and their flipper lengths center near 217 mm, both substantially higher than the other species. The boxplots also show relatively tight distributions, suggesting that Gentoos are not only larger on average but also more uniformly large.

Adelie and Chinstrap penguins overlap more closely in size, but some distinctions still emerge. Chinstraps tend to have slightly higher mean values in both body mass and flipper length—about 3733 g vs. 3701 g in mass and 195.8 mm vs. 190.0 mm in flipper length. The boxplots reflect this modest separation: Chinstraps show a somewhat larger median and a slightly higher overall range in both measurements. However, the overlap between these two species indicates that size alone may not reliably differentiate them in the field.

Overall, the results show a pattern where Gentoo penguins are consistently the largest in both body mass and flipper length, while Adelie and Chinstrap penguins are more similar, with Chinstraps being marginally larger on average. These size differences may reflect ecological distinctions among the species, such as diet, foraging range, and habitat use.


###   How do bill length and bill depth interact to help differentiate penguin species? Can these measurements be used to identify species differences effectively?
```{r}
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point(size = 3, alpha = 0.8) +
  labs(title = "Bill Length vs Bill Depth by Species",
       x = "Bill Length (mm)",
       y = "Bill Depth (mm)") +
  theme_minimal()

```
The scatterplot shows that bill length and bill depth vary systematically across the three penguin species, and these differences form distinct clusters. Because each species occupies its own region in the bill-length/bill-depth space, the two measurements together provide a strong basis for distinguishing among the species.

Adelie penguins tend to have shorter bills but relatively deep ones. Their points cluster on the left side of the plot with higher bill-depth values. In contrast, Gentoo penguins generally have much longer bills but are characterized by shallower depths, forming a cluster on the right with lower vertical values. Chinstrap penguins fall between the two—typically with moderate-to-long bill lengths and moderate-to-high bill depths—creating their own, somewhat overlapping but still visually separable group.

These patterns indicate that bill length alone or bill depth alone may not cleanly separate the species, but the interaction of the two measurements allows for clear differentiation. By examining where a penguin falls in relation to these clusters, one can effectively predict its species. This makes bill dimensions not just biologically meaningful traits but also practical tools for classification.


### Is there a relationship between flipper length and body mass across species? Do larger flippers indicate heavier penguins, and does this relationship differ by species?
```{r}
library(ggplot2)
library(dplyr)

penguins_clean |> 
  group_by(species) |> 
  summarize(correlation = cor(flipper_length_mm, body_mass_g))

ggplot(penguins_clean, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point() +
  labs(
    title = "Relationship Between Flipper Length and Body Mass by Species",
    x = "Flipper Length (mm)",
    y = "Body Mass (g)",
    color = "Species"
  )
```

The scatterplot reveals a strong positive relationship between flipper length and body mass across all penguin species. In general, penguins with longer flippers tend to be heavier, and this trend is consistent regardless of species. However, the strength and position of this relationship differ by species, which is further supported by the correlation values.

Gentoo penguins (blue points) clearly occupy the upper-right range of the plot, indicating that they have both the longest flippers (around 210–230 mm) and the highest body mass (4500–6000 g). Their correlation coefficient (r ≈ 0.71) is the strongest among the three species, highlighting a very pronounced relationship between flipper size and overall body mass. This suggests that Gentoos are significantly larger overall compared to the other species.

Chinstrap penguins (green points) fall in the middle range, with moderate flipper lengths and body mass values. Their trend line shows a positive association supported by a moderately strong correlation (r ≈ 0.64). While not as steep or wide as the Gentoo group, the Chinstrap data still demonstrates a clear connection between flipper length and body mass.

Adelie penguins (red points) cluster in the lower-left region, indicating shorter flippers and lower body mass. Their positive relationship is visible but weaker compared to the other species, which aligns with their lower correlation value (r ≈ 0.46). This indicates that while flipper length still relates to body mass for Adelies, the relationship is less strong.

Overall, the plot and correlation analysis together indicate that larger flippers correspond to heavier body mass, but the magnitude of this relationship varies by species. Gentoo penguins exhibit the strongest and most pronounced trend, while Adelie and Chinstrap penguins show smaller ranges of body size. This suggests that differences in body size are both species-specific and biologically meaningful.

After seeing a positive trend in the scatterplot, I became curious whether the relationship looked the same within each species. To check this, I calculated the average body mass for groups of flipper lengths:

```{r}
penguins_clean$flipper_group <- cut(penguins_clean$flipper_length_mm,
                                    breaks = 4)

tapply(penguins_clean$body_mass_g,
       list(penguins_clean$species, penguins_clean$flipper_group),
       mean, na.rm = TRUE)
```

These grouped averages increased steadily within each species, reinforcing the visual pattern that longer flippers generally correspond to heavier penguins. Even though each species has different absolute sizes, the upward pattern appeared in all three species, confirming the overall positive association.

To test whether the trend was possibly driven by a few extreme values, I examined the minimum and maximum values:
```{r}
tapply(penguins_clean$body_mass_g, penguins_clean$species, range)
tapply(penguins_clean$flipper_length_mm, penguins_clean$species, range)
```

These ranges showed that Gentoo penguins have much larger bodies overall, raising skepticism that part of the trend might simply reflect species-level differences. Because each species occupies a different size range, the overall positive trend could be partly due to species clustering rather than a universal relationship within all penguins. This made me cautious about assuming the trend applies equally to every individual penguin.

### How does sex influence body size? Are male penguins consistently larger than female penguins across all species, and does this vary in strength across difference species?
```{r}
penguins_clean |> 
  group_by(species, sex) |> 
  summarize(mean_mass = mean(body_mass_g, na.rm = TRUE))

ggplot(penguins_clean, aes(x = sex, y = body_mass_g, fill = sex)) +
  geom_boxplot(alpha = 0.8) +
  facet_wrap(~ species) +
  labs(
    title = "Body Mass by Sex Across Penguin Species",
    x = "Sex",
    y = "Body Mass (g)"
  ) 
```

The boxplot shows that male penguins are consistently larger in body mass than female penguins across all three species, indicating a clear pattern of sexual dimorphism in the Palmer Penguins dataset. However, the magnitude of this size difference varies between species, which is further supported by the mean body mass values.

For Adelie penguins, the difference between males and females is noticeable but modest. Male Adelies have an average body mass of about 4043 g, while females average around 3369 g, placing both groups within the 3200–4200 g range. Although their distributions overlap slightly, the numerical means confirm a moderate sex-based size difference.

In Chinstrap penguins, the size difference between males and females becomes more pronounced. Male Chinstraps average approximately 3939 g, compared to 3527 g for females. This 400-gram gap, paired with the reduced overlap in the boxplots, indicates stronger separation than observed in Adelies.

The strongest sex difference appears in Gentoo penguins. Males are significantly heavier, with an average mass of about 5485 g, while females average around 4680 g. This substantial difference of more than 800 grams, along with minimal overlap in the boxplots, highlights the most pronounced sex-based size difference across the three species.

Overall, the plot and mean mass values together demonstrate that sex influences body size in all species, but the strength of this effect increases from Adelie to Chinstrap to Gentoo. This suggests that sexual dimorphism is a consistent biological pattern in penguins, though more pronounced in some species than others.

To explore whether the differences were consistent or if a few unusual values might be influencing the result, I looked at the standard deviations:

```{r}
tapply(penguins_clean$body_mass_g,
       list(penguins_clean$species, penguins_clean$sex),
       sd, na.rm = TRUE)
```

Gentoo males had a standard deviation of about 313 g, noticeably larger than Adelie females (≈269 g) or Chinstrap females (≈285 g), which shows that body mass varies a lot within this group. This suggested to me that the difference between male and female Gentoo penguins may not be as uniform as the boxplot makes it appear.

### Does the island a penguins inhabits affect its body measurements? Are physical differences across islands still present even after accounting for species?

```{r}

library(palmerpenguins)
library(dplyr)
library(ggplot2)


penguins_clean <- penguins %>%
drop_na()


head(penguins_clean)

```


```{r}

penguins_clean %>%
group_by(island) %>%
summarize(
mean_mass = mean(body_mass_g, na.rm = TRUE),
mean_flipper = mean(flipper_length_mm, na.rm = TRUE),
mean_bill_length = mean(bill_length_mm, na.rm = TRUE),
mean_bill_depth = mean(bill_depth_mm, na.rm = TRUE)
)

```



```{r}

ggplot(penguins_clean, aes(x = island, y = body_mass_g, fill = island)) +
geom_boxplot(alpha = 0.7) +
labs(title = "Body Mass by Island", y = "Body Mass (g)", x = "Island") +
theme_minimal()

ggplot(penguins_clean, aes(x = island, y = flipper_length_mm, fill = island)) +
geom_boxplot(alpha = 0.7) +
labs(title = "Flipper Length by Island", y = "Flipper Length (mm)", x = "Island") +
theme_minimal()

```

The first boxplot produced shows the distribution of penguin body mass by island. Biscoe island has the highest median body mass with around 4,750 grams, and its moderate spread suggests a fairly wide range of body sizes among the penguins living there. Torgersen Island has a lower median body mass of around 3,500 grams, with a spread similar to Dream Island, indicating moderate variability. Overall, these patterns indicate that penguins on Biscoe tend to be heavier, and those on Dream and Torgersen tend to be lighter. There is also the occasional outlier, such as the one on Dream Island. However, the boxes and whiskers overlap considerably across the three islands, suggesting that island alone does not create large or distinct differences in body mass. 

Body mass does vary slightly by island, but because the distributions overlap so much, these differences likely reflect the species that live on each island rather than the island itself. Larger species such as Gentoos are more common on some islands, which likely explains the higher body masses seen on Biscoe Island.


The second boxplot produced shows the distribution of penguin flipper length by island. The boxplots for flipper length also show clear but modest differences across islands. Penguins on Biscoe Island have the highest median flipper length at around 215 millimeters. Dream Island penguins show a median flipper length of around 195 millimeters that falls between Biscoe Island and Torgersen Island. Penguins on Torgersen Island have the smallest median flipper length at around 190 millimeters, although there is a higher outlier. These patterns show that Biscoe Island penguins generally have the longest flippers, Torgersen Island penguins the shortest, and Dream Island penguins fall in the middle.

As with body mass, flipper length differences across islands appear small and slightly overlapping, indicating that the island itself has only a modest influence. Differences in species composition across islands likely explain much of the variation, as larger species like Gentoo are more common on certain islands.




```{r}

ggplot(penguins_clean, aes(x = island, y = body_mass_g, fill = island)) +
geom_boxplot(alpha = 0.7) +
facet_wrap(~ species) +
labs(title = "Body Mass by Island Within Each Species", y = "Body Mass (g)", x = "Island") +
theme_minimal()

ggplot(penguins_clean, aes(x = island, y = flipper_length_mm, fill = island)) +
geom_boxplot(alpha = 0.7) +
facet_wrap(~ species) +
labs(title = "Flipper Length by Island Within Each Species", y = "Flipper Length (mm)", x = "Island") +
theme_minimal()

```




The first faceted boxplot shows the distribution of body mass for Adelie, Chinstrap, and Gentoo penguins across Biscoe, Dream, and Torgersen Islands. For Adelie penguins, body mass is fairly consistent across islands, with Dream Island showing a slightly lower median and moderate spread. Chinstrap penguins appear only on Dream Island on this plot with two outliers and a small spread. Biscoe appears to have a slightly higher spread, indicating more variability in body mass on that island. Gentoo penguins display the largest body masses, particularly on Biscoe Island, with a higher median and a wider spread, while Dream Island has slightly lower medians and Torgersen has the fewest data points, suggesting less representation.

Overall, within each species, body mass differences between islands are relatively small, with overlap in some cases. This indicates that species identity is the primary driver of body mass variation, while the island a penguin inhabits has only a modest influence. Larger medians or spreads on certain islands are likely due to the presence of more individuals from larger species (Gentoo) rather than a strong island effect.

While some small variations exist across islands, body mass is largely determined by species, and any influence of the island itself is minor.



The second faceted boxplot shows the distribution of flipper length for Adelie, Chinstrap, and Gentoo penguins across Biscoe, Dream, and Torgersen Islands. For Adelie penguins, median flipper length is within the same range (190) for each island, although there are a few outliers. Chinstrap penguins on Dream Island tend to have a slighty higher median (197). While Gentoo penguins on Biscoe island have the highest median flipper length (217).

Overall, within each species, flipper length differences across islands are relatively modest, with some overlap between distributions. This suggests that species identity is the main factor influencing flipper size, while the island a penguin inhabits has only a minor effect. Slightly higher medians or greater spreads on certain islands are likely due to the presence of more individuals from larger species, such as Gentoo, rather than a strong island-specific influence.

Flipper length is primarily determined by species, with island location playing only a subtle, secondary role.





### Which variable (bill length, bill depth, flipper length, or body mass) provides the strongest prediction of species classification? How do combines variables compare to single-variable predication?


```{r}
library(ggplot2)
library(dplyr)
library(palmerpenguins)

penguins_clean <- penguins |>
  drop_na(species, bill_length_mm, bill_depth_mm,
          flipper_length_mm, body_mass_g)


ggplot(penguins_clean, aes(species, bill_length_mm, fill = species)) +
  geom_boxplot() +
  labs(title = "Bill Length by Species",
       x = "Species", y = "Bill Length (mm)") +
  theme_minimal()


ggplot(penguins_clean, aes(species, bill_depth_mm, fill = species)) +
  geom_boxplot() +
  labs(title = "Bill Depth by Species",
       x = "Species", y = "Bill Depth (mm)") +
  theme_minimal()


ggplot(penguins_clean, aes(species, flipper_length_mm, fill = species)) +
  geom_boxplot() +
  labs(title = "Flipper Length by Species",
       x = "Species", y = "Flipper Length (mm)") +
  theme_minimal()


ggplot(penguins_clean, aes(species, body_mass_g, fill = species)) +
  geom_boxplot() +
  labs(title = "Body Mass by Species",
       x = "Species", y = "Body Mass (g)") +
  theme_minimal()


```


The boxplots show differences in all four measurements across Adelie, Chinstrap, and Gentoo penguins. Gentoo penguins stand out as having both the longest flippers and the highest body mass, making these two variables the strongest indicators of species identity. Flipper length, in particular, shows almost no overlap between Gentoo and the other species, suggesting it is the single most distinguishable trait. Bill length varies across the three species, with Chinstrap penguins having the longest bills, Adelie the shortest, and Gentoo falling in between. Bill depth also varies, as Chinstrap penguins tend to have a larger bill depth, although Adelie penguins aren't close behind.

These patterns indicate that while all four traits vary by species, flipper length and body mass provide the clearest overall separation, specifically in Gentoo penguins, whereas bill depth and bill length are most useful for distinguishing Chinstrap penguins. 

Combining all four variables further improves species classification, resolving cases where a single variable might overlap between species. Combining all four variables further improves species classification, resolving cases where a single variable might overlap between species, and allows for nearly perfect prediction of species identity, showing that using multiple measurements provides the most reliable differentiation.

## Conclusion

Our analysis of the Palmer Penguins dataset successfully met the goals outlined in our introduction. By examining key body size measurements across species, sex, and island, we identified clear and consistent patterns in physical traits. Species differences emerged as the strongest influence on body size, with Gentoo penguins generally showing the largest measurements and Adelie and Chinstrap following behind. Sex also played a role, with males tending to be slightly larger than females across species. In contrast, the island where penguins were found showed only minor variation, suggesting that location has far less impact on size than species or sex.

These findings support our initial goal of understanding how biological and environmental factors relate to variation in penguin traits. While environmental differences between islands appear modest, species-level differences were distinct and meaningful. Looking ahead, further research using larger datasets or additional ecological variables such as diet, temperature, or nesting conditions, could help clarify why these physical differences exist and how they relate to penguin survival in their habitats.
