---
title: "Analysis on Physical Characteristics of Penguins in the Palmer Archipelago"
author: "Jyotika Sharma, Selena Cooper, Grace Kolker"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The goal of this project is to explore the Palmer Penguins dataset to better understand the differences in physical traits between penguin species. This dataset includes measurements collected on three penguins species (Adelie, Gentoo, and Chinstrap) living on multiple islands in the Palmer Archipelago in Antarctica. By analyzing these physical measurements, we hope to identify patterns, species differneces, and relationships between body size variables. Understanding these traits can help explain how environmental conditions and biological factors may influence species variation and survival.

In pursuit of the stated goal, we will explore the following questions:

1. How do body size measurements vary between penguin species? Are certain species consistently larger in one or more measurements such as body mass or flipper length? - Selena

2. How do bill length and bill depth interact to help differentiate penguin species? Can these measurements be used to identify species differences effectively? - Selena

3. Is there a relationship between flipper length and body mass across species? Do larger flippers indicate heavier penguins, and does this relationship differ by species? - Jyotika

4. How does sex influence body size? Are male penguins consistently larger than female penguins across all species, and does this vary in strength across difference species? - Jyotika 

5. Does the island a penguins inhabits affect its body measurements? Are physical differences across islands still present even after accounting for species? - Grace

6. Which variable (bill length, bill depth, flipper length, or body mass) provides the strongest prediction of species classification? How do combines variables compare to single-variable predication? - Grace 

These are the main questions we intend to answer through analysis of the Palmer Penguins dataset. With the findings, we will be able to draw meaningful conclusions about physical trait differences and the ecological factors that influence species variation in Antarctic penguins.

## Data

### Structure

The link to the dataset is
```{r}
library(palmerpenguins)
data(package = "palmerpenguins")
```

### Cleaning
```{r}
library(tidyverse)

penguins_clean <- penguins |> 
  drop_na()

# Check dimensions before/after
dim(penguins)
dim(penguins_clean)

```
The Palmer Penguins dataset was already very clean and well-structured upon import, containing properly formatted variable types and no obvious outliers or inconsistent entries. We did not need to convert sex to factor since it already is well. The only issues present were several missing values in the bill measurements, flipper length, body mass, and sex variables. To ensure accurate and consistent analysis across all variables, we removed rows containing any missing values using drop_na(). This reduced the dataset from 344 observations to 333 complete cases.

### Variables
```{r}
dim(penguins)
str(penguins)
```
There are 344 penguins in the dataset. There are 3 different species of penguins collected from 3 different islands in the Palmer Archipelago.

* species: the biological species of each penguin (Adelie, Gentoo, and Chinstrap)
* island: The island within the Palmer Archipelago where the penguin was observed (Biscoe, Dream, and Torgersen)
* bill_length_mm: The length of the penguin's bill (from tip to base) measured in milimeters
* bill_depth_mm: Thickness (depth) fo the penguin's bill measured at the base in milimeters
* flipper_length_mm: Length of the penguin's flipper from shoulder to tip measured in milimeters
* body_mass_g: Body weight of the penguin measured in grams
* sex: Sex of the penguin (male or female)
* year: The year the penguin was sampled during the Palmer Station field seasons (2007, 2008, 2009)

## Results

###  How do body size measurements vary between penguin species? Are certain species consistently larger in one or more measurements such as body mass or flipper length?

### How do bill length and bill depth interact to help differentiate penguin species? Can these measurements be used to identify species differences effectively?

### Is there a relationship between flipper length and body mass across species? Do larger flippers indicate heavier penguins, and does this relationship differ by species?
```{r}
library(ggplot2)
library(dplyr)

penguins_clean |> 
  group_by(species) |> 
  summarize(correlation = cor(flipper_length_mm, body_mass_g))

ggplot(penguins_clean, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point() +
  labs(
    title = "Relationship Between Flipper Length and Body Mass by Species",
    x = "Flipper Length (mm)",
    y = "Body Mass (g)",
    color = "Species"
  )
```

The scatterplot reveals a strong positive relationship between flipper length and body mass across all penguin species. In general, penguins with longer flippers tend to be heavier, and this trend is consistent regardless of species. However, the strength and position of this relationship differ by species, which is further supported by the correlation values.

Gentoo penguins (blue points) clearly occupy the upper-right range of the plot, indicating that they have both the longest flippers (around 210–230 mm) and the highest body mass (4500–6000 g). Their correlation coefficient (r ≈ 0.71) is the strongest among the three species, highlighting a very pronounced relationship between flipper size and overall body mass. This suggests that Gentoos are significantly larger overall compared to the other species.

Chinstrap penguins (green points) fall in the middle range, with moderate flipper lengths and body mass values. Their trend line shows a positive association supported by a moderately strong correlation (r ≈ 0.64). While not as steep or wide as the Gentoo group, the Chinstrap data still demonstrates a clear connection between flipper length and body mass.

Adelie penguins (red points) cluster in the lower-left region, indicating shorter flippers and lower body mass. Their positive relationship is visible but weaker compared to the other species, which aligns with their lower correlation value (r ≈ 0.46). This indicates that while flipper length still relates to body mass for Adelies, the relationship is less strong.

Overall, the plot and correlation analysis together indicate that larger flippers correspond to heavier body mass, but the magnitude of this relationship varies by species. Gentoo penguins exhibit the strongest and most pronounced trend, while Adelie and Chinstrap penguins show smaller ranges of body size. This suggests that differences in body size are both species-specific and biologically meaningful.

After seeing a positive trend in the scatterplot, I became curious whether the relationship looked the same within each species. To check this, I calculated the average body mass for groups of flipper lengths:

```{r}
penguins_clean$flipper_group <- cut(penguins_clean$flipper_length_mm,
                                    breaks = 4)

tapply(penguins_clean$body_mass_g,
       list(penguins_clean$species, penguins_clean$flipper_group),
       mean, na.rm = TRUE)
```

These grouped averages increased steadily within each species, reinforcing the visual pattern that longer flippers generally correspond to heavier penguins. Even though each species has different absolute sizes, the upward pattern appeared in all three species, confirming the overall positive association.

To test whether the trend was possibly driven by a few extreme values, I examined the minimum and maximum values:
```{r}
tapply(penguins_clean$body_mass_g, penguins_clean$species, range)
tapply(penguins_clean$flipper_length_mm, penguins_clean$species, range)
```

These ranges showed that Gentoo penguins have much larger bodies overall, raising skepticism that part of the trend might simply reflect species-level differences. Because each species occupies a different size range, the overall positive trend could be partly due to species clustering rather than a universal relationship within all penguins. This made me cautious about assuming the trend applies equally to every individual penguin.

### How does sex influence body size? Are male penguins consistently larger than female penguins across all species, and does this vary in strength across difference species?
```{r}
penguins_clean |> 
  group_by(species, sex) |> 
  summarize(mean_mass = mean(body_mass_g, na.rm = TRUE))

ggplot(penguins_clean, aes(x = sex, y = body_mass_g, fill = sex)) +
  geom_boxplot(alpha = 0.8) +
  facet_wrap(~ species) +
  labs(
    title = "Body Mass by Sex Across Penguin Species",
    x = "Sex",
    y = "Body Mass (g)"
  ) 
```

The boxplot shows that male penguins are consistently larger in body mass than female penguins across all three species, indicating a clear pattern of sexual dimorphism in the Palmer Penguins dataset. However, the magnitude of this size difference varies between species, which is further supported by the mean body mass values.

For Adelie penguins, the difference between males and females is noticeable but modest. Male Adelies have an average body mass of about 4043 g, while females average around 3369 g, placing both groups within the 3200–4200 g range. Although their distributions overlap slightly, the numerical means confirm a moderate sex-based size difference.

In Chinstrap penguins, the size difference between males and females becomes more pronounced. Male Chinstraps average approximately 3939 g, compared to 3527 g for females. This 400-gram gap, paired with the reduced overlap in the boxplots, indicates stronger separation than observed in Adelies.

The strongest sex difference appears in Gentoo penguins. Males are significantly heavier, with an average mass of about 5485 g, while females average around 4680 g. This substantial difference of more than 800 grams, along with minimal overlap in the boxplots, highlights the most pronounced sex-based size difference across the three species.

Overall, the plot and mean mass values together demonstrate that sex influences body size in all species, but the strength of this effect increases from Adelie to Chinstrap to Gentoo. This suggests that sexual dimorphism is a consistent biological pattern in penguins, though more pronounced in some species than others.

To explore whether the differences were consistent or if a few unusual values might be influencing the result, I looked at the standard deviations:

```{r}
tapply(penguins_clean$body_mass_g,
       list(penguins_clean$species, penguins_clean$sex),
       sd, na.rm = TRUE)
```

Gentoo males had a standard deviation of about 313 g, noticeably larger than Adelie females (≈269 g) or Chinstrap females (≈285 g), which shows that body mass varies a lot within this group. This suggested to me that the difference between male and female Gentoo penguins may not be as uniform as the boxplot makes it appear.

### Does the island a penguins inhabits affect its body measurements? Are physical differences across islands still present even after accounting for species?

### Which variable (bill length, bill depth, flipper length, or body mass) provides the strongest prediction of species classification? How do combines variables compare to single-variable predication?

## Conclusion



