---
title: "Analysis on Physical Characteristics of Penguins in the Palmer Archipelago"
output: html_document
date: "2025-11-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Analysis on Physical Characteristics of Penguins in the Palmer Archipelago

#### Jyotika Sharma, Selena Cooper, Grace Kolker

## Introduction

The goal of this project is to explore the Palmer Penguins dataset to better understand the differences in physical traits between penguin species. This dataset includes measurements collected on three penguins species (Adelie, Gentoo, and Chinstrap) living on multiple islands in the Palmer Archipelago in Antarctica. By analyzing these physical measurements, we hope to identify patterns, species differneces, and relationships between body size variables. Understanding these traits can help explain how environmental conditions and biological factors may influence species variation and survival.

In pursuit of the stated goal, we will explore the following questions:

1. How do body size measurements vary between penguin species? Are certain species consistently larger in one or more measurements such as body mass or flipper length?

2. How do bill length and bill depth interact to help differentiate penguin species? Can these measurements be used to dientify species differences effectively?

3. Is there a relationship between flipper length and body mass across species? Do larger flippers indicate heavier penguins, and does this relationship differ by species?

4. How does sex influecne body size? Are male penguins consistently larger than female penguins across all species, and does this vary in strnegth across difference species?

5. Does the isalnd a penguins inhabits affect its body measurements? Are physcial differences across islands still present even after accoutning for species?

6. Which variable (bill length, bill depth, flipper length, or body mass) provides the strongest prediction of speceis classification? How do combines variables compare to single-variable predication?

These are the main questions we intend to answer through analysis of the Palmer Penguins dataset. With the findings, we will be able to draw meaningful conclusions about physical trait differences and the ecological factors that influence species variation in Antarctic penguins.

## Data

### Structure

The link to the dataset is
```{r}
library(palmerpenguins)
data(package = "palmerpenguins")
```

### Cleaning

### Variables

## Results
